---
title: "BSMM-lab-3"
subtitle: "BSMM 8740 Fall 2023"
author: "Tavish Bhandari"
date: "October 12, 2023"
format: html
editor: visual
self-contained: true
---

## Setup

Load packages and data:

```{r load-pkg-data}
#| message: false
boston_cocktails <- readr::read_csv('data/boston_cocktails.csv')
```

## Exercises

### Exercise 1

The median measure amount across across all cocktails is 1.

```{r}
library(magrittr)   # the pipe
library(tidyverse)  # for data wrangling + visualization
library(tidymodels) # for modeling
library(gt)         # for pretty tables
# Use introduce to explore variables in the dataset 
DataExplorer::introduce(boston_cocktails)
# Calculate the median measure number
median_measure_number <- median(boston_cocktails$measure_number)
print(median_measure_number)
```

### Exercise 2

The **Leap Frog Highball** contains 2 of gin

```{r}
library(tidyverse)
#install.packages("janitor")
library(janitor)
#Select the columns mentioned
select_columns <- boston_cocktails |>
  select(name, category, ingredient, measure_number)
# Pivot the table to create columns for each ingredient
pivoted_table <- select_columns |>
  pivot_wider(names_from = ingredient, values_from = measure_number, values_fill = 0)
# Clean column names
cleaned_table <- pivoted_table |>
  janitor::clean_names()
# Evaludate how much gin
gin_in_leap_frog_highball <- cleaned_table |>
  filter(name == "Leap Frog Highball") |>
  select(gin)
print(gin_in_leap_frog_highball)
```

### Exercise 3

0 predictor variables are prepped by the recipe.

```{r}
library(recipes)
#View(boston_cocktails)
# Create a recipe object using the loaded dataset
recipe_obj <- recipe(~ ., data = boston_cocktails) |>
  update_role(name, category, new_role = "id") |>
  step_dummy(all_nominal()) |>
  step_normalize(all_numeric()) |>
  step_pca(all_numeric(), num_comp = 5)
# Prepare the data
prepped_data <- prep(recipe_obj)
# Count the number of predictor variables prepped by the recipe
num_predictor_vars <- length(prepped_data$predictors)
print(num_predictor_vars)
```

### Exercise 4

On average the most used ingredient in the Boston Cocktails dataset is cranberry juice.

```{r}
# Summarize the data to find the most used ingredient on average
ingredient_summary <- boston_cocktails |>
  group_by(ingredient) |>
  summarize(avg_measure_number = mean(measure_number, na.rm = TRUE)) |>
  arrange(desc(avg_measure_number))
most_used_ingredient <- ingredient_summary$ingredient[1]
print(most_used_ingredient )
```

### Exercise 5

Describe describe the drinks represented by PC1?

Row_id : 0 and 0.25, it has a positive loading in PC1, then when row_id is in this range, it contributes positively to higher values of PC1.

Measure_number: 0 and -0.7, it has a negative loading in PC1, then when measure_number is in this range, it contributes negatively to higher values of PC1.

Ingredient_number: 0 and 0.7, it has a positive loading in PC1, then when ingredient_number is in this range, it contributes positively to higher values of PC1.

```{r}
library(recipes)
library(dplyr)
library(forcats)
library(ggplot2)
numeric_columns <- select_if(boston_cocktails, is.numeric)
boston_cocktails_recipe <-
  recipe(~., data = numeric_columns) %>% 
  update_role(., row_id, ingredient_number, measure_number) %>% 
  step_naomit(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()
boston_cocktails_pca <- 
  boston_cocktails_recipe %>% 
  tidy(id = "pca", matrix = "X") # Use matrix = "X" to keep the original data
# Filter for components PC1 to PC5 and mutate them as factors
boston_cocktails_pca_filtered <- boston_cocktails_pca %>%
  filter(component %in% c("PC1", "PC2", "PC3", "PC4", "PC5")) %>%
  mutate(component = fct_inorder(component))
# Create the PCA plot
ggplot(boston_cocktails_pca_filtered, aes(x = value, y = terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, nrow = 1) +
  labs(y = NULL) +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 14, face = "bold"))
```

### Exercise 6

The characteristic alcoholic beverage of each of the first 4 principle components is less than or equal to 0.

```{r}
# Load required libraries
library(dplyr)
library(gt)
# Create a function to color cells based on value
color_cells <- function(x) {
  ifelse(x < 0, "red", "green")
}
# Slice the top 8 ingredients by component based on absolute value
top_ingredients_table <- boston_cocktails_pca_filtered %>%
  filter(component %in% c("PC1", "PC2", "PC3", "PC4","PC5")) %>%
  group_by(component) %>%
  slice_max(order_by = abs(value), n = 8) %>%
  ungroup() %>%
  pivot_wider(names_from = component, values_from = terms)
# Modify the table to add cell background colors using gt
for (col in names(top_ingredients_table)[-1]) {
  top_ingredients_table[[col]] <- sapply(top_ingredients_table[[col]], function(x) {
    cell_style <- color_cells(x)
    sprintf('%s', cell_style, x)
  })
}
# Create the gt table
table_pca_ingredients <- top_ingredients_table %>%
  gt() %>%
  tab_style(
    style = cell_fill(
      color = color_cells(0)
    ),
    locations = cells_body()
  )
# Print the table
table_pca_ingredients
```

### Exercise 7

How would you interpret the results of a PCA analysis for a client?

In this PCA analysis, PC1 and PC2 fail to distinctly separate data points, suggesting a complex or low-variance structure in the dataset. Further exploration of higher-dimensional components may be needed to uncover hidden patterns.

```{r}
library(dplyr)
library(recipes)
library(ggplot2)
# Create the PCA recipe
rec <- recipe(~., data = boston_cocktails)
pca_trans <- rec %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3)
pca_estimates <- prep(pca_trans, training = boston_cocktails)
pca_data <- bake(pca_estimates, boston_cocktails)
# Extend the range for the plot
rng <- extendrange(c(pca_data$PC1, pca_data$PC2))
# Create PCA with threshold
with_thresh <- rec %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), threshold = 0.99)
with_thresh <- prep(with_thresh, training = boston_cocktails)
baked_with_thresh <- bake(with_thresh, boston_cocktails)
# Print tidy PCA results
tidy(pca_trans, number = 2)
tidy(pca_estimates, number = 2)
# Create the scatter plot of PC1 and PC2 with labels
ggplot(pca_data, aes(PC1, PC2, label = name)) +
  geom_point(aes(color = category), alpha = 0.7, size = 2) +
  geom_text(check_overlap = TRUE, hjust = "inward") + 
  labs(color = NULL)
```
